Tails or Types? A Critical Systematic Review of Taxometric Studies in Neurodevelopmental Disorders

### 2025-07-23 Paper/Paper.qmd
update paper

---
title: "Tails or Types? A Critical Systematic Review of Taxometric Studies in Neurodevelopmental Disorders"
shorttitle: "Taxometric Review Neurodevelopmental Disorders"
authors:
  - name: Enrico Toffalini
    corresponding: true
    orcid: 0000-0002-1404-5133
    email: enrico.toffalini@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Margherita Calderan
    corresponding: false
    orcid: 0009-0005-5668-5162
    email: margherita.calderan@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Riccardo Pagan
    corresponding: false
    email: riccardo.pagan.3@studenti.unipd.it
    affiliation: Department of Social and Developmental Psychology, University of Padua, Italy
  - name: Valentina Tobia
    corresponding: false
    email: tobia.valentina@unisr.it
    affiliation: University Vita-Salute San Raffaele, Milan, Italy
author-note:
  disclosures:
    study-registration: 'The review was preregistered at https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0'
    financial-support: 'Financial support: Project 2022KBW99S "Tails or types? Testing the dimensional hypothesis in neurodevelopmental disorders", funded by Next Generation EU, Mission 4, Componente 1, CUP C53D23004210006'
    gratitude: "We are grateful to Athanasia Vasileiadou who first pointed us to some of the reviewed articles."
    data-sharing: null
    related-report: null
    conflict-of-interest: null
    authorship-agreements: null
abstract: "..."
keywords: [Neurodevelopmental disorders, Taxometric analysis, Autism, ADHD, Learning disorders]
bibliography: references.bib
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-typst: default
  apaquarto-pdf:
    documentmode: man
---

In recent years, research on neurodevelopmental disorders (NDDs) has undergone widespread reconsiderations. Traditional diagnostic categories defined in the DSM-5 [@dsm5], have been increasingly called into question in favor of a dimensional and transdiagnostic view [e.g., @astle2022trandiagnostic; @michelini2024world; @happe2020looking; @happe2021both]. The DSM-5 itself has incorporated dimensional aspects, for example in defining a broad Autistic Spectrum Disorder (ASD) condition instead of a set of different autism-related diagnoses. Several arguments have contributed to this: the large heterogeneity within diagnostic groups, the poor alignment between data-driven clusters and DSM-based categories, the frequent comorbidity and phenotypic correlations across conditions, and the idea that individuals with NDDs might represent, at least in part, the tails of developmental continua, that are quantitatively rather than qualitatively different from the so-called neurotypical population [@michelini2024world; @happe2021both; @mammarella2021nocore].

Clarifying whether NDDs truly represent categories, rather than just tails of continuous distributions that span across the population, is important for different reasons. These involve quests about social and personal self-identity by involved individuals, discussions about diagnostic framing, comorbidity, and the choice on interventions in clinical settings, the structuring of classes, schools, and educational methods when dealing with neurodiversity, and applications to research design, planning of studies, and power analysis in research settings [@caviola2024editorial]. But most importantly, this is relevant for our understanding of the ontological status of reality, which should matter for scientific endeavor.

A dimensional approach fits well with long-standing principles in the study of individual differences in psychology, and has long characterized research on intelligence and personality [@caviola2024editorial]. It also seems appropriate for NDDs given accumulating evidence of polygenicity and multifactorial etiology [@plomin2005generalist; @kovas2006generalist; @pennington2012multiple; @astle2020core; @caviola2024editorial; @demontis2023adhdgenome]. In fact, most behavioral traits are believed to be influenced not by one or a few, but by many small, additive causes [@chabris2015fourth]. According to a famous principle known in statistics as the central limit theorem [@fischer2011central], when many individual causes (even discrete ones, e.g., gene variants) add up they converge to a joint distribution that tends to continuity and normality: this is, in fact, the "secret" of the Gaussian distribution. Nonetheless, a principled categorical view still resists in neurodevelopmental research, for example concerning ASD [e.g., @happe2021both; @chown2021bit; @frazier2023asd], subtypes within NDDs [e.g., @burgess2018taxodyslexia] or the possible emergence of new data-driven taxa [@astle2022trandiagnostic].

Adjudicating between whether a clinical diagnostic category is characterized by an underlying structure that is categorical or dimensional is precisely the focus of taxometric analysis, a set of quantitative methods that have been developed since the 1990s [e.g., @meehl1995taxometric]. Since then, taxometric analysis has been widely applied in psychopathology, generally providing evidence that most mental health conditions (e.g., depression, anxiety) are best conceptualized as dimensional rather than categorical [@haslam2012review; @haslam2020metaanalysis]. In brief, taxometric methods such as MAMBAC, MAXCOV, MAXEIG, and L-Mode, leverage patterns of covariation among observed indicators to test alternative competing latent models that are categorical or dimensional [@ruscio2007taxometric; @ruscio2004checklist; @cole2004taxometric]. Such methods have become increasingly refined through simulation-based techniques such as the Comparison Curve Fit Index (CCFI), which provide criteria for objective decision-making [@ruscio2018using].

Despite this, the use of taxometric analysis in neurodevelopmental research has been limited. Most existing studies focus on adult psychopathology, with few exceptions [@beauchaine2003taxometrics; @haslam2020metaanalysis]. @happe2021both and @chown2021bit in their debate on whether ASD is a category or a continuum not even mention the potential of taxometric analysis. In their recent influential article, @astle2022trandiagnostic briefly presents taxometry but gives it little emphasis, prevailing reviewing studies that employ cluster analysis. However, the limitations of clustering for taxometric purposes have been discussed for decades. Clustering algorithms tend to impose (rather than seek) categorical structure, often fail to recover known taxa in simulations, and generally provide weak tests of taxonicity [@beauchaine2003taxometrics]. Recent studies simulating realistic psychological data, characterized by a combination of (even modestly) skewed and correlated indicators, further suggest that common clustering algorithms easily detect pseudoclusters while having low power to test the existence of true taxa [@toffalini2022entia; @toffalini2024clusters]. Indeed, @astle2022trandiagnostic interpret results from clustering studies as evidence that data-driven structures fail to align with the expected DSM-5 categories, rather than as proper tests of taxonicity.

Taxometric analysis is not without limitations, however, some of which are shared with cluster analysis. Commonly cited threats to validity of taxometric conclusions include poor indicator validity, nuisance covariance, low base rate, and artificial admixture \[...\]. Poor indicator validity occurs when taxometric analysis is run on variables that present little mean differences between individuals with and without the target disorder (i.e., putatively, the taxon and the complement). Generally, it is recommended that Cohen's d ≫ 1 on each indicator. Nuisance covariance means that there is strong residual correlations after accounting for taxa (i.e., both within the taxon and the complement), thus reflecting the use of redundant indicators; residual r \> 0.40 is often regarded as problematic. Low base rate means that individuals with the taxonic condition are too few in the sample for taxometric analysis to correctly identify them; n \< 30 is often regarded as problematic. This is obviously a problem when the diagnostic category is rare (e.g., 1-2% of the population). Finally, artificial admixture consists of mixing participants selected or recruited using alternate criteria. This is often (incorrectly) used to compensate for the low base rate, with many individuals with a diagnosis being recruited separately from the rest of the sample to increase the number of observations with the putative taxon. Poor indicator validity and low base rate risk masking detection of taxa, while strong nuisance covariance and artificial admixture risk to detecting pseudotaxa by mimicking categorical structures.

@@

The main goal of this systematic review is to critically examine the use of taxometric methods in the study of neurodevelopmental disorders (NDDs) within the published literature. We focused our review on the most commonly studied NDD categories: Autism Spectrum Disorder (ASD), Attention-Deficit/Hyperactivity Disorder (ADHD), and all forms of Specific Learning Disorders (SLDs), along with the closely related condition of Language Disorder/Impairment (the latter was not included in the preregistered protocol and thus constitutes a deviation). We excluded Intellectual Disability, as its etiological basis is often well identified and taxonic by definition in many cases (e.g., Down syndrome), or associated with identifiable medical risk factors. Residual non-syndromic cases, in turn, are thought to reflect the lower tail of general mental ability, and thus as inherently dimensional.

A core aim of the review is to critically assess the methodological quality and credibility of findings reported in published taxometric studies. We evaluated potential sources of bias such as artificial admixture, and whether sufficiently complete methodological information was reported to judge other threats to validity, including low indicator validity, nuisance covariance, and the skewness of indicators. (The latter is relevant when scales assess rare or extreme behaviors, and can create the illusion of categorical structure).

Following this systematic review, we present a Monte Carlo simulation study to empirically demonstrate the extent to which artificial admixture can lead to false detection of categorical latent structures in dimensional data, both when using taxometric analysis and cluster analysis.

@@

The systematic review was conducted in accordance with the above listed goals, the preregistration protocol, and the PRISMA guidelines. The methods are detailed below.

@@

A systematic search was conducted on four databases: PsycINFO, PubMed, Scopus, Web of Science. No time limits were set, however virtually no records from before 1990 were expected, as taxometric methods were not yet developed or diffuse then. The following search query was performed on titles, abstracts, and/or keywords:

((“taxometric\*” OR “MAMBAC” OR “MAXCOV” OR “MAXEIG” OR “L-Mode”) AND (“autis\*” OR “ASD” OR “attention-deficit\*” OR “ADHD” OR “learning dis\*” OR “reading dis\*” OR “dyslexia” OR “math\* dis\*" OR "math\* diff\*"))

Where the wildcard "\*" was not supported (i.e., PubMed) it was replaced with full terms ("disorder", "disability" instead of "dis\*"; "mathematical", "mathematics", "math" instead of "math\*").

Language disorder was subsequently added as a deviation from preregistration, replacing the second part of the above search query with ("language imp\*" OR "language dis\*"). In addition, the full search query was launched on Google Scholar and the first 10 pages were screened; the references of and citations to @haslam2020metaanalysis were searched on Google Scholar for additional results; and a deep research via OpenAI's gpt-4o was launched providing it the preregistration protocol as context. All promising records thus emerging and not previously identified via databases, were listed under the *"Identification of studies via other methods"* section of the PRISMA flowchart.

@@

Inclusion criteria:

-   The study applies at least one taxometric procedure (e.g., MAMBAC, MAXCOV, L-Mode, MAXEIG);

-   The target population is individuals with NDDs or related traits;

-   The study draws a conclusion about the latent structure (categorical vs. dimensional) or reports sufficient information for an independent reader to reach a conclusion based on those results (e.g., reports CCFIs).

Exclusion criteria:

-   Theoretical, methodological, or simulation articles without empirical data;

-   Articles not reporting any behavioral data (e.g., only neuroimaging or neurophysiological data);

-   Reviews and/or meta-analyses;

-   Studies using other latent structure methods only (e.g., LPA, mixture modeling) without any taxometric methods being employed.

@@

Two independent human coders (MC and RP) screened and coded all articles. All discrepancies were resolved through discussion, and later revised by the PI (ET). No AI tools were used in the process, except in occasional cases by the human coders to assist them clarify the content of articles.

Since a single article could report multiple (sub)samples and/or multiple taxometric analyses on the same sample, the full coded dataframe followed a long-form data structure, with some articles repeating over multiple rows. Coded variables are listed in the data dictionary table reported in the preregistration protocol, to which a few variables for notes and coders agreement were added.

@@

@@

We focus on artificial admixture because it is a very powerful threat to validity that could relatively easily be avoided by design when planning a study, and because as suggested by the result of the systematic review presented below, it is unfortunately frequent. There are different types of admixtures; @ruscio2004checklist discusses three cases. First, as mentioned above, the situation where clinical and nonclinical cases are separately recruited and merged. Second, the case of splitting a sample into subsamples (using alternative criteria) and then using those to test different taxometric questions. Third, trimming away observations of the putative complement (i.e., individuals without the disorder) to increase the base rate. All of these procedures artificially impose categorical structures to the data that do not exist in the population. It is worth mentioning that artificial admixture is also a threat for any type of cluster analysis, for the exact same reasons that it is a problem for taxometric analysis.

We simulate a dimensional scenario that makes it visually compelling how artificial admixture of the first type mentioned above creates pseudotaxa. A single normally-distributed latent variable $X$ is generated, that defines the target condition. A series of observed indicators $x1, x2, etc.$ are simulated, incorporating the $X$ latent variable plus some random normally-distributed noise ($R^2 = 0.77$). Individuals above the 99^th^ percentile of $X$ receive a diagnosis via clinical assessment. A sample of $N = 1000$ individuals is collected from the general population. Since the base rate is low (on average, only about $n = 10$ cases are expected to naturally present a diagnosis), researchers select and recruit an additional $n = 1000$ cases with a clinical diagnosis from the general population, thus creating artificial admixture. All data simulation is available in the online materials at xxxxxxxxxxxxxxxx.

An instance of the simulation, plotting two observed variables for illustrative purposes, is shown in @FIGURE_xx. Color is added for further clarity to distinguish individuals with and without a diagnosis, but two clusters still clearly emerge. Indicator validity is good (about $Cohen's$ $d = 2$), while nuisance covariance is partly problematic: for the clinical cases $r \approx 0$, but for the nonclinical cases $r = 0.61$. It is worth reminding that no taxa exist here, and the appearance of a categorical structure is mainly imposed by admixture, as shown below.

>>>>

---
title: "Tails or Types? A Critical Systematic Review of Taxometric Studies in Neurodevelopmental Disorders"
shorttitle: "Taxometric Review Neurodevelopmental Disorders"
authors:
  - name: Enrico Toffalini
    corresponding: true
    orcid: 0000-0002-1404-5133
    email: enrico.toffalini@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Margherita Calderan
    corresponding: false
    orcid: 0009-0005-5668-5162
    email: margherita.calderan@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Riccardo Pagan
    corresponding: false
    email: riccardo.pagan.3@studenti.unipd.it
    affiliation: Department of Social and Developmental Psychology, University of Padua, Italy
  - name: Valentina Tobia
    corresponding: false
    email: tobia.valentina@unisr.it
    affiliation: University Vita-Salute San Raffaele, Milan, Italy
author-note:
  disclosures:
    study-registration: 'The review was preregistered at https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0'
    financial-support: 'Financial support: Project 2022KBW99S "Tails or types? Testing the dimensional hypothesis in neurodevelopmental disorders", funded by Next Generation EU, Mission 4, Componente 1, CUP C53D23004210006'
    gratitude: "We are grateful to Athanasia Vasileiadou who first pointed us to some of the reviewed articles."
    data-sharing: https://osf.io/ys5ad/?view_only=0e2812acfabf43fc8af01f883c7331ba
    related-report: null
    conflict-of-interest: null
    authorship-agreements: null
abstract: "Taxometric methods are designed to test whether latent structures underlying psychological constructs are categorical (taxonic) or dimensional. Despite their relevance to the nature of neurodevelopmental disorders (NDDs), their application has been limited and often methodologically flawed. We conducted a preregistered systematic review to critically evaluate this literature. The review identified all published taxometric studies on autism, ADHD, learning, and language disorders, with a focus on methodological biases such as artificial admixture. While studies on ADHD and SLI supported dimensional structures, most studies on autism and NDD subtypes reported taxonic findings, often in the presence of methodological threats. A Monte Carlo simulation illustrates how artificial admixture can spuriously generate taxonic results in otherwise dimensional data. We conclude that evidence for categorical models in NDDs is frequently undermined by methodological artifacts, reinforcing the need for caution, but also advocating for more, high-quality taxometric studies in the field of NDDs."
keywords: [Neurodevelopmental disorders, Taxometric analysis, Autism, ADHD, Learning disorders]
bibliography: references.bib
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-typst: default
  apaquarto-pdf:
    documentmode: man
---

```{r,message=F,warning=F}
library(readxl)
library(kableExtra)
library(stringr)
library(dplyr)
```

In recent years, research on neurodevelopmental disorders (NDDs) has undergone significant reconsideration. Traditional diagnostic categories defined in the DSM-5 [@dsm5] have increasingly been questioned in favor of a dimensional and transdiagnostic perspective [e.g., @astle2022trandiagnostic; @michelini2024world; @happe2020looking; @happe2021both; @sonuga2020kuhnian; @caviola2024editorial]. Notably, the DSM-5 itself has incorporated some dimensional features (for example by defining a broad Autism Spectrum Disorder [ASD] in place of distinct diagnoses). Several factors have motivated this shift: the marked heterogeneity within diagnostic groups, the weak correspondence between data-driven clusters and DSM-based categories, the frequent comorbidity and phenotypic correlations across conditions, and the view that individuals with NDDs may represent, at least in part, the tails of developmental continua, quantitatively rather than qualitatively different from the so-called neurotypical population [@michelini2024world; @happe2021both; @mammarella2021nocore].

Determining whether NDDs truly represent categories, rather than just tails of continuous distributions spanning the general population, is important for several reasons, including questions of social and personal identity for affected individuals, implications for diagnostic framing, comorbidity, and intervention choices in clinical contexts, and consequences for the organization of classrooms, educational practices, and even the design and power analysis of research studies [@caviola2024editorial]. But more fundamentally, this question is relevant for our understanding of the ontological status of psychological phenomena.

A dimensional approach aligns well with long-standing principles in the study of individual differences in psychology, and has long characterized research on intelligence and personality [@caviola2024editorial]. It also seems well-suited for NDDs, given a growing body of evidence supporting polygenicity and multifactorial etiology [@plomin2005generalist; @kovas2006generalist; @pennington2012multiple; @astle2020core; @caviola2024editorial; @demontis2023adhdgenome]. Most behavioral traits are understood to result not from one or a few causes, but from the additive effects of many small influences [@chabris2015fourth]. The central limit theorem [@fischer2011central] posits that, when many individual causes (such as discrete genetic variants) contribute additively, the resulting distribution tends to continuity and normality (this is, in fact, the "secret" of the Gaussian distribution). Nonetheless, a principled categorical view still resists in neurodevelopmental research, for example concerning ASD [e.g., @happe2021both; @chown2021bit; @frazier2023asd], subtypes within NDDs [e.g., @burgess2018taxodyslexia] or the proposed emergence of data-driven taxa [@astle2022trandiagnostic].

Taxometric analysis is a set of quantitative methods originally introduced by Paul Meehl and colleagues in the 1970s and refined over decades [e.g., @meehl1995taxometric], explicitly aimed at establishing whether a clinical disorder is characterized by an underlying categorical or dimensional structure. It has been widely applied in psychopathology, and typically suggested that most mental health conditions (e.g., depression, anxiety) are better conceptualized as dimensional [@haslam2012review; @haslam2020metaanalysis]. In brief, taxometric procedures, such as MAMBAC, MAXCOV, MAXEIG, and L-Mode examine covariation patterns among observed indicators to compare competing latent models: one categorical, the other dimensional [@ruscio2007taxometric; @ruscio2004checklist; @cole2004taxometric]. These methods now feature simulation-based tools like the Comparison Curve Fit Index (CCFI), which provide a quantitative measures to facilitate inference [@ruscio2018using].

Unfortunately, the use of taxometric analysis has remained limited in neurodevelopmental research, which most existing studies focusing on adult psychopathology [@beauchaine2003taxometrics; @haslam2020metaanalysis]. In their debate on whether ASD is best conceptualized as a category or a continuum, @happe2021both and @chown2021bit do not even mention such methods. Similarly, in their influential review, @astle2022trandiagnostic briefly acknowledge taxometry, but puts much more emphasis on studies using cluster analysis. However, the limitations of clustering methods for assessing taxonicity have been discussed for decades: clustering algorithms tend to impose (rather than test for) categorical structures, they provide weak tests of taxonicity, and often fail to recover known taxa in simulations [@beauchaine2003taxometrics; @ruscio2004boundaryissues]. Recent simulations using psychologically realistic data (featuring both skewness and intercorrelated indicators) suggest that commonly used clustering methods are prone to detecting pseudoclusters while presenting low power to identify true taxa [@toffalini2022entia; @toffalini2024clusters]. Indeed, @astle2022trandiagnostic interpret clustering results as evidence that data-driven groups do not align with DSM-5 categories, rather than as compelling tests of latent structural form.

Taxometric analysis is not without limitations, some of which are shared with cluster analysis. Commonly cited threats to validity of conclusions include poor indicator validity, nuisance covariance, low base rate, and artificial admixture [@cole2004taxometric; @haslam2020metaanalysis; @ruscio2004checklist; @ruscio2018using].

Poor indicator validity implies that indicators present little mean difference between individuals with and without the target condition (i.e., the putative taxon and complement). Ideally, Cohen’s *d* for each indicator should substantially exceed 1 (*d* $\gg$ 1). Nuisance covariance refers to strong residual correlations among indicators after accounting for group membership (indicating redundancy among variables). Low base rate means that too few individuals in the sample belong to the taxon, making accurate detection difficult (samples with *n* < 30 for the taxonic group are usually deemed insufficient). This is particularly relevant when studying rare diagnostic categories. Artificial admixture occurs when participants are combined from different recruitment sources or selection criteria, often as an attempt to compensate for a low base rate. For instance, clinical cases may be recruited separately from controls and merged into a single sample. In general, poor indicator validity and low base rate reduce power to detect true taxa, while high nuisance covariance and artificial admixture increase the risk of detecting pseudotaxa by mimicking a categorical structure.

@@

The primary aim of this systematic review is to critically examine the use of taxometric methods in studies of NDDs within the published literature. We focused on the most commonly investigated NDD categories: Autism Spectrum Disorder (ASD), Attention-Deficit/Hyperactivity Disorder (ADHD), and all forms of Specific Learning Disorders (SLDs), along with the closely related condition of Language Disorder/Impairment. (Language disorders were not included in the preregistered protocol and therefore constitute a minor deviation.) We excluded Intellectual Disability, as its etiology is often clearly identifiable and taxonic/syndromic by definition in many cases (e.g., Down syndrome), or linked to specific medical risk factors, while residual non-syndromic cases are typically regarded as reflecting the lower tail of general mental ability and thus inherently dimensional.

A crucial goal was to assess the methodological quality and credibility of taxometric findings. We specifically evaluated potential sources of bias, such as artificial admixture, and assessed whether studies provided sufficient methodological information to evaluate other known threats to validity, including low indicator validity, high nuisance covariance, and skewed indicator distributions. (The latter is particularly relevant when scales assess rare or extreme behaviors, which can create the illusion of a categorical structure.)

Following the review, we report a Monte Carlo simulation study designed to empirically demonstrate the extent to which artificial admixture can lead to the spurious detection of categorical latent structures in otherwise dimensional data, using both taxometric and clustering methods.

@@

The systematic review was conducted in accordance with the Preferred Reporting items for Systematic Reviews and Meta-Analysis (PRISMA) 2020 guidelines. A review protocol was preregistered on OSF registries at the following link: https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0 The methods are detailed below.

@@

A systematic search for relevant peer-reviewed articles was conducted in April 2025 using four electronic databases: PsycINFO, PubMed, Scopus, and Web of Science. No time limits were applied, although records predating 1990 were not expected, as taxometric methods were not yet widely adopted at that time. The following search query—reflecting the methods and target populations of interest—was applied to titles, abstracts, and keywords:

```
(("taxometric*" OR "MAMBAC" OR "MAXCOV" OR "MAXEIG" OR "L-Mode") AND 
("autis*" OR "ASD" OR "attention-deficit*" OR "ADHD" OR "learning dis*" 
OR "reading dis*" OR "dyslexia" OR "math* dis*" OR "math* diff*"))
```

In databases where the wildcard `*` was not supported (e.g., PubMed), it was replaced with full terms (e.g., `"disorder"`, `"disability"` instead of `"dis*"`; `"mathematical"`, `"mathematics"`, `"math"` instead of `"math*"`).

Language disorder was subsequently added as a deviation from the preregistered protocol, replacing the second part of the search query with:

```
("language imp*" OR "language dis*")
```

Additionally, the full search query was launched on Google Scholar, and the first 10 pages of results were screened manually. References cited in, and articles citing, @haslam2020metaanalysis were also screened via Google Scholar. Finally, an exploratory search using OpenAI’s GPT-4o was conducted, using the preregistration protocol as contextual input. All promising records identified through these additional methods and not retrieved via database searches were listed under the *"Identification of studies via other methods"* section of the PRISMA flowchart.

@@

Inclusion criteria were as follows:  
1. The study applied at least one taxometric procedure (e.g., MAMBAC, MAXCOV, L-Mode, MAXEIG);  
2. The target population consisted of individuals with NDDs or related traits;  
3. The study drew a conclusion about the latent structure (categorical vs. dimensional), or reported sufficient information (e.g., CCFIs) for an independent reader to infer such a conclusion.

Exclusion criteria were:  
1. The article was theoretical, methodological, or a simulation study without empirical data;  
2. The study did not report any behavioral data (e.g., only neuroimaging or neurophysiological data; note: this criterion was later reconsidered, but no studies were excluded for this reason in practice);  
3. The study was a review and/or meta-analysis;  
4. The study employed latent structure methods other than taxometric analysis (e.g., latent profile analysis, mixture modeling), with no taxometric procedures reported.

In addition, consistent with the preregistered protocol, the search was limited to published, peer-reviewed journal articles. This decision reflected the primary aim of providing a critical review of the existing empirical literature. However, one potentially relevant non-journal record was identified and, while excluded from the systematic review, is discussed separately in the *Discussion* section.

@@

Two independent human coders (MC and RP) screened and coded all articles. No AI tools were involved in the coding process, except in occasional cases where coders used them to help clarify specific article content. All discrepancies were resolved through discussion and later reviewed by the principal investigator (ET).

Since a single article could report multiple (sub)samples and/or multiple taxometric analyses on the same sample, the full coded dataset followed a long-form data structure, with some articles appearing across multiple rows. Coded variables were defined in the data dictionary included in the preregistration protocol, with a few additional fields added to capture coder notes and agreement metrics.

Inter-coder agreement was fair to good across all phases. In the title and abstract screening phase, Cohen’s *κ* = 0.79 (34 out of 38 independent decisions matched). In the full-text eligibility screening, Cohen’s *κ* = 0.52 (15 out of 19). For the final data extraction, 84% of rows (71 out of 85) were extracted by both coders. Agreement on critical variables was high: for the "authors' taxonic conclusion" agreement was 94%; for "artificial admixture", it was 95% at the sample or subsample level.

@@

A total of 110 records were identified through database searches. After removing 67 duplicates and excluding 5 additional records (dissertations or book chapters, only one of which appeared to include primary taxometric analyses but could not be retrieved), 38 unique articles were screened at the title and abstract level.

Nineteen records were excluded at this stage. Full-text assessment was then conducted on the remaining 19 articles, of which 7 were excluded for not meeting the inclusion criteria: 4 did not apply taxometric methods, and 3 did not target neurodevelopmental disorders.

An additional 4 records were identified through citation tracking and online browsing. Of these, 2 were excluded (1 was a methodological article, and 1 did not employ taxometric analysis).

In total, 14 studies met inclusion criteria and were included in the review. The PRISMA flowchart in @fig-prisma summarizes the study selection process.

@fig-prisma *about here*

![PRISMA 2020 flow diagram illustrating the study screening and selection process](Figures/PRISMA_flow_diagram.png){#fig-prisma}

```{r}
df = data.frame(read_excel("../Literature-review/data/data_extraction.xlsx"))
```

A summary of the included studies, simplified to one row per article, is presented in @tbl-summary. A full reporting of all coded data for each individual taxometric analysis is available in the Supplemental Online Material. The total number of taxometric analyses was `r nrow(df)`, ranging from `r min(table(df$ID_article))` to `r max(table(df$ID_article))` analyses per article. The targeted disorders were ADHD (6 articles), ASD (6 articles), dyslexia (1 article), and Specific Language Impairment (SLI; 2 articles), with one article addressing both ADHD and ASD. Most articles (10 out of 14) examined whether a given disorder constituted a taxon, while the remaining 4 focused on whether putative subtypes within a disorder exhibited taxonic structure.

The most commonly used taxometric method was MAMBAC, applied in 13 out of 14 articles and in `r sum(grepl("MAMBAC", df$Taxometric_methods_used))` out of `r nrow(df)` total analyses. This was followed by MAXEIG (9 articles), L-Mode (7 articles), MAXCOV (2 articles), and MAXSLOPE (1 article). Most studies employed more than one taxometric procedure, with a median of `r median(str_count(df$Taxometric_methods_used, ";") + 1)` methods per analysis. In 11 out of 14 articles, interpretation was based on a quantitative fit index (Fit~d~ or, in more recent studies, the CCFI), while the remaining 3 relied on visual inspection of taxometric curves. The number of indicators entered into analyses ranged from `r min(df$Indicators_number, na.rm = TRUE)` to `r max(df$Indicators_number, na.rm = TRUE)`, with a median of `r median(df$Indicators_number, na.rm = TRUE)` (excluding one article in which the number was not clearly reported). The median sample size across all analyses was `r median(df$Sample_size)`, or `r median((df |> group_by(ID_article) |> summarize(avg = mean(Sample_size)))$avg)` when averaged within articles.

Taxonic conclusions were mixed. All five articles targeting ADHD as a disorder concluded in favor of a dimensional structure, as did both articles focusing on SLI. In contrast, all four articles examining ASD as a disorder leaned toward, or explicitly supported, a categorical (taxonic) structure. The remaining four articles investigated subtypes of disorders, and all leaned toward taxonic conclusions—either strongly or predominantly [@stevens2018functional on ADHD subtypes; @munson2008asd on ASD subtypes; @obrien2012dyslexia on dyslexia subtypes], or partially [@ingram2008defining, which supported ASD subtypes as taxa depending on the set of indicators used]. In total, 8 out of 14 articles reached a taxonic conclusion in at least some of their taxometric analyses. Specifically, `r sum(df$Taxonic_conclusion)` out of `r nrow(df)` total analyses—representing `r round(mean(df$Taxonic_conclusion) * 100, 1)`%—yielded taxonic results. These were primarily associated with ASD: `r sum(grepl("ASD", df$Target_disorder) & df$Taxonic_conclusion == 1)` out of `r sum(grepl("ASD", df$Target_disorder))` analyses involving ASD (i.e., `r round(100 * sum(grepl("ASD", df$Target_disorder) & df$Taxonic_conclusion == 1) / sum(grepl("ASD", df$Target_disorder)), 1)`%) supported a categorical structure.

Artificial admixture was evident or strongly suspected in at least part of the analyses in 8 out of 14 articles, involving a total of `r sum(df$Artificial_admixture)` out of `r nrow(df)` analyses. Notably, articles that incurred artificial admixture tended to align with those that reached taxonic conclusions: 6 of the 8 articles with suspected admixture also reported taxonic findings. This association was even more pronounced at the level of individual analyses. A taxonic conclusion was reached in `r sum(df$Taxonic_conclusion[df$Artificial_admixture == 1])` out of `r sum(df$Artificial_admixture == 1)` analyses (`r round(100 * mean(df$Taxonic_conclusion[df$Artificial_admixture == 1]), 1)`%) where admixture was evident or suspected. In contrast, only `r sum(df$Taxonic_conclusion[df$Artificial_admixture == 0])` out of `r sum(df$Artificial_admixture == 0)` analyses (`r round(100 * mean(df$Taxonic_conclusion[df$Artificial_admixture == 0]), 1)`%) reached a taxonic conclusion when admixture was likely or certainly absent.

Remaining methodological issues concerned missing information about key descriptive statistics and validity indices. In particular, **nuisance covariance**—that is, correlations among indicators within diagnostic groups or within the putative taxon and complement—could be coded for only 2 out of 14 articles. Where reported, it was small to negligible in @james2016latent, but exceeded 0.40 for many indicators in @frazier2023asd; notably, both studies reached a taxonic conclusion. **Skewness** of indicator score distributions was reported in only 5 articles. When available, skewness values were often near or beyond ±1 for at least one indicator in roughly one-third of analyses. However, there was no clear association between skewness and whether a taxonic conclusion was reached. By contrast, **indicator validity** was more consistently reported, and could be coded in all but 3 articles. Where available, validity (typically expressed as Cohen’s *d*) was almost always large or very large (*d* $\gg$ 1), consistent with best-practice recommendations for taxometric analysis.

@tbl-summary *about here*

```{r}
#| label: tbl-summary
#| tbl-cap: Overview summary of the reviewed articles."
summary_tab = read_excel("Tables/Review-Summary.xlsx",sheet=2)
kable(summary_tab, format = "latex", booktabs = TRUE) |>
  kable_styling(font_size = 8, latex_options = c("striped", "scale_down")) |>
  column_spec(1:ncol(summary_tab),width="5em") |>
  landscape()
```

## Additional Use of Cluster Analysis

Among the 14 identified articles, 6 also employed clustering methods—typically latent class analysis (LCA) or latent profile analysis (LPA). In 5 of these 6 cases, clustering was conducted in parallel with taxometric analysis [@james2016latent; @munson2008asd; @frazier2023asd; @frazier2007adhd; @deserno2023probing], while in one case it was presented as a secondary analysis [@frazier2010asd]. All six studies targeted ADHD and/or ASD, and in all cases clustering results were interpreted as supporting taxonic conclusions. These interpretations generally converged with those from taxometric analysis, with one exception: in @frazier2007adhd, clustering results were partially inconclusive but suggested the presence of subgroups within ADHD, thus contrasting with the dimensional structure favored by the taxometric analysis. In all but two studies [@frazier2007adhd; @munson2008asd], we identified at least some risk of artificial admixture.

@@

We focus on artificial admixture because it represents a powerful threat to validity that is both common (as suggested by our review) and avoidable by design through better planning of a study. Different types of admixture can be incurred. @ruscio2004checklist describes three main forms. First (and most frequently) is the case where clinical and nonclinical participants are separately recruited and merged into a single sample. Second, admixture can occur when a sample is artificially split into subsamples based on different selection criteria, and these are then used to test different taxometric questions [cf. "subtractive" compound sampling in @haslam2020metaanalysis]. Third, researchers may trim observations from the putative complement (i.e., individuals without the disorder) to inflate the base rate of the taxon. Each of these procedures risks imposing a categorical structure onto data that might be dimensional in nature. It is worth noting that artificial admixture poses a threat not only to taxometric analysis, but also to clustering methods, for the same reasons: both infer latent structure based on observed covariation, which can be distorted by selection artifacts.

```{r}
load("../R/simulateAdmixtResults.RData")
```

We simulate a dimensional scenario to visually demonstrate how artificial admixture, specifically of the first type described above, can produce pseudotaxonic results. A single normally distributed latent variable $X$ is generated to represent the underlying clinical trait of interest. A set of observed indicators ($x_1$, $x_2$, $x3$) are then simulated as linear functions of $X$ plus normally distributed random noise, yielding an $R^2$ of 0.77 for each indicator. Individuals above the 99th percentile of $X$ are assumed to receive a diagnosis via global clinical assessment. A sample of $N = 1000$ individuals is drawn from the general population. Given the low base rate, only about $n = 10$ cases are expected to naturally receive a diagnosis. To address this, researchers draw a second sample consisting of an additional $n = 1000$, targeting clinically diagnosed cases, and merge it with the original sample, thereby creating an artificial admixture. The full simulation code and data are available in the online materials:  
<https://osf.io/ys5ad/?view_only=0e2812acfabf43fc8af01f883c7331ba>

An instance of the simulation, illustrating two observed variables $x_1$ and $x_2$, is presented in @fig-admixture. Color is used to distinguish individuals with and without a diagnosis, but even without this aid, two clusters visually emerge. This apparent separation is misleading: no latent categories exist in the data. The appearance of a categorical structure is a consequence of artificial admixture. In this simulated example, indicator validity is high (Cohen’s *d* ≈ 2), suggesting strong separation between groups on each variable. However, nuisance covariance is partly problematic: among diagnosed individuals, the correlation between indicators is close to zero ($r \approx 0.05$), whereas among non-diagnosed individuals, it is substantial ($r \approx 0.60$). This asymmetry, combined with differential recruitment, contributes to the illusion of a latent taxon.

@fig-admixture *about here*
